---
title: 'Calculus: Differentiation and Its Application'
subtitle: A statistical perspective
format:
  ubd40-revealjs:
    slide-level: 2
    transition: fade
    auto-stretch: false
    width: 1250  # 1050
    height: 760  # 700
    self-contained: false
    chalkboard: true
    toc: false
    toc-depth: 1
    multiplex: true
    code-block-height: 700px
    # html-table-processing: none
author:
  - name: Dr. Haziq Jamil
    orcid: 0000-0003-3298-1010
    # email: haziq.jamil@ubd.edu.bn | `https://haziqj.ml/aiti-talk`
    affiliations: 
      - 'Assistant Professor in Statistics, Universiti Brunei Darussalam'
      - '<span style="font-style:normal;">[`https://haziqj.ml/uitm-calculus/`](https://haziqj.ml/uitm-calculus/)</span>'
date: 2025-06-14
# bibliography: refs.bib
execute:
  echo: false
  freeze: auto
  cache: false
---

## (Almost) Everything you ought to know...

### ...about calculus in the first year



Let $f:\mathcal X \to \mathbb R$ be a *real-valued* function defined on an input set $\mathcal X$.

::: {.fragment}

::: {#def-derivatives}

#### Differentiability

$f(x)$ is said to be *differentiable* at a point $x \in \mathcal X$ if the limit 

$$
L = \lim_{h \to 0} \frac{f(x + h) - f(x)}{h}
$${#eq-def-derivative}
exists.
If $L$ exists, we denote it by $f'(x)$ or $\frac{df}{dx}(x)$, and call it the *derivative* of $f$ at $x$.
Further, $f$ is said to be differentiable on $\mathcal X$ if it is differentiable at every point in $\mathcal X$.
:::

:::

::: {.fragment}
For now, we assume $\mathcal X \subseteq \mathbb R$, and will extend to higher dimensions later.
:::

## Some examples


::: {.columns}

::: {.column width="50%"}
<br>

| Function | Derivative |
| --- | --- |
| $f(x) = x^2$ | $f'(x) = 2x$ |
| $f(x) = \sum_{n} a_n x^n$ | $f'(x) = \sum_{n} n a_n x^{n-1}$ | 
| $f(x) = \sin(x)$ | $f'(x) = \cos(x)$ |
| $f(x) = \cos(x)$ | $f'(x) = -\sin(x)$ |
| $f(x) = e^x$ | $f'(x) = e^x$ |
| $f(x) = \ln(x)$ | $f'(x) = \frac{1}{x}$ | 
| | |
: {.table-ubd}
:::

::: {.column width="5%"}
:::

::: {.column width="45%"}

<br>

We can derive it "by hand" using the definition.
Let $f(x) = x^2$.
Then,

<br>
<br>

$$
\begin{align}
\lim_{h \to 0} & \frac{f(x + h) - f(x)}{h} \\
&= \lim_{h \to 0} \frac{(x + h)^2 - x^2}{h} \\
&= \lim_{h \to 0} \frac{x^2 + 2xh + h^2 - x^2}{h} \\[0.5em]
&= \lim_{h \to 0} 2x + h \\[0.5em]
&= 2x.
\end{align}
$$

:::

:::

::: aside
[Taylor series](https://en.wikipedia.org/wiki/Taylor_series) is a powerful tool for approximating functions using derivatives. 
:::


## Graphically...

```{=html}
<script src="https://www.desmos.com/api/v1.11/calculator.js?apiKey=dcb31709b452b1cf9dc26972add0fda6"></script>

<div id="desmos-derivative" style="position: relative; width: 100%; height:88%;"></div>

<script>
  const elt = document.getElementById("desmos-derivative");
  const calculator = Desmos.GraphingCalculator(elt, {
    expressions: true,
    settingsMenu: false,
    zoomButtons: true
  });

  calculator.setExpressions([
    { id: 'f', latex: 'f(x) = x^2' },
    { id: 'a', latex: 'a = 1', sliderBounds: { min: -1, max: 3 } },
    { id: 'h', latex: 'h = 1.5', sliderBounds: { min: 0.001, max: 1.5 } },
    { id: 'pt1', latex: '(a, f(a))', showLabel: true },
    { id: 'pt2', latex: '(a + h, f(a + h))', showLabel: true },
    { id: 's', latex: 'L = \\frac{f(a+h) - f(a)}{h}' },
    { id: 'tangent', latex: 'y = L(x - a) + f(a)', color: Desmos.Colors.RED }
  ]);
  
  calculator.setMathBounds({
    left: -1,
    right: 3,
    bottom: -0.5,
    top: 8
  });

</script>
```

## What is a derivative?

The derivative of a function tells you:    

- 🚀 How fast the function is _changing_ at any point  
- 📐 The **slope** of the tangent line at that point

```{r}
#| fig-height: 2.8
#| fig-width: 7
#| out-width: 100%
library(tidyverse)

f <- function(x) -x^2
f_prime <- function(x) -2*x
x_pts <- c(-1, 0, 1)
fty <- factor(c("f increasing", "f stationary", "f decreasing"),
                levels = c("f increasing", "f stationary", "f decreasing"))

curve_df <- tibble(x = seq(-2, 2, length.out = 200), y = f(x))

tangent_df <- lapply(x_pts, function(a) {
  tibble(x = seq(a - 1, a + 1, length.out = 100),
         y = f_prime(a) * x + f(a) - f_prime(a) * a,
         x0 = a,
         type = fty[match(a, x_pts)])
}) |> bind_rows()

point_df <- tibble(x = x_pts, y = f(x_pts), type = fty)

ggplot() +
  geom_line(data = curve_df, aes(x, y), color = "black", linewidth = 0.8) +
  geom_line(data = tangent_df, aes(x, y), color = "red3", linewidth = 0.8) +
  geom_point(data = point_df, aes(x, y), color = "steelblue3", size = 2) +
  facet_wrap(~ type) +
  labs(x = "x", y = "f(x)") +
  theme_bw()
```

## The concept of optimisation

- When $f$ is some kind of a "reward" function, then the value of $x$ that maximises $f$ is highly of interest.

- Derivatives help us find maxima (or minima): Solve $f'(x) = 0$.

- Example: Find the maximum of $f(x) = -3x^4 + 4x^3 + 12x^2$.

## Graphically...

```{r}
#| fig-height: 4.1
#| fig-width: 7
#| out-width: 100%
library(ggrepel)

f <- function(x) -3 * x ^ 4 + 4 * x ^ 3 + 12 * x ^ 2#-x^4 + 2*x^3 + 3*x^2 - 2*x
x_crit <- c(-1, 0, 2)
df <- tibble(x = seq(-2, 3, length.out = 500)) %>%
  mutate(y = f(x))

critical_points <- tibble(
  x = x_crit,
  label = c("Local Max", "Local Min", "Global Max"),
  color = RColorBrewer::brewer.pal(3, "Set1")
) %>%
  mutate(y = f(x))

shade_df <- tibble(
  xmin = c(-Inf, x_crit[1], x_crit[2]),
  xmax = c(x_crit[1], x_crit[2], x_crit[3]),
  fill = rep(c("grey75", "white"), length.out = 3)
)

pm_df <- tibble(
  x = ((c(-2.2, x_crit, 3.2) + c(x_crit, 3.2, 0)) / 2)[1:4],
  y = 35
)

ggplot(df, aes(x, y)) +
  geom_rect(data = shade_df, aes(xmin = xmin, xmax = xmax, ymin = -Inf, 
                                 ymax = Inf, fill = fill), 
            inherit.aes = FALSE, alpha = 0.4) +  
  scale_fill_identity() +
  geom_label(data = pm_df, aes(x, y, label = c("+", "−", "+", "−")),
            size = 8, color = "black", vjust = -0.5) +
  geom_line(color = "black", size = 1) +
  geom_point(data = critical_points, aes(x, y, color = label), size = 3,
             show.legend = FALSE) +
  geom_text_repel(
    data = critical_points, 
    aes(x, y, label = label, color = label), 
    size = 4.5, 
    seed = 4, 
    #nudge_y = 0.5,
    box.padding = 1, 
    point.padding = 0.4, 
    show.legend = FALSE
  ) +
  scale_colour_brewer(palette = "Set1") +
  coord_cartesian(ylim = c(-8, 43)) +
  labs(x = "x", y = "f(x)") +
  theme_bw()
```


## How do we know if it's a maxima or minima?

Second derivative test: Measure the **change in slope** around the critical point $\hat x$, i.e. $f''(\hat x) = \frac{d}{dx}\left( \frac{df}{dx}(x) \right)$.



| Behaviour of $f$ near $\hat x$               | $f''(\hat x)$       | Shape             | Conclusion     |
|-------------------------------|---------------------|----------------------|-------------------|----------------|
| Increasing → Decreasing       | $f''(\hat x) < 0$ | Concave (∩)  | Local maximum  |
| Decreasing → Increasing       | $f''(\hat x) > 0$ | Convex (∪)    | Local minimum  |
| No sign change / flat region  |  $f''(\hat x) = 0$   | Unknown / flat    | Inconclusive   |


## Uses of calculus in everyday life

1. **Physics**: Calculating velocity and acceleration from position functions.
2. **Economics**: Finding marginal cost and revenue functions.
3. **Engineering**: Optimizing designs for strength and efficiency.
4. **Biology**: Modeling population growth rates.
5. **Finance**: Analyzing risk and return in investment portfolios.



## Motivation

Show students how theoretical tools they’re learning now are used to derive estimators;


Plant the idea that Statistics isn’t just “data” or “Excel”, but has real mathematical depth;
Give a clear reason to care about derivatives, gradients, Hessians, and Jacobians in practice.

## Functions



## Derivatives



## Derivatives

Goal: Shift mindset from high-school calculus to tools for optimization and multidimensional problems.
	•	Quick recap: derivative as slope, second derivative as curvature
	•	Extension to multivariable functions:
	•	Partial derivatives
	•	Gradient as direction of steepest ascent
	•	Hessian as local curvature matrix
	•	Applications in optimization: what it means to maximize or minimize a function with multiple variables

Example: A simple multivariate function $f(x, y) = -x^2 - y^2 + 4x + 6y$.
Find its critical point using partial derivatives, use the Hessian to classify.


## Maximum likelihood estimation

- Define likelihood: $L(\theta) = \prod_{i=1}^n f(x_i \mid \theta)$
  - Log-likelihood: easier to differentiate
  - MLE found by solving $\frac{d}{d\theta} \ell(\theta) = 0$

## Examples

Emphasize: the maximum is found because second derivative is negative (concavity)

## Fisher information 

Second derivative = Fisher Information (intuition)

- Negative expected second derivative of log-likelihood
- Measures how peaked the likelihood is → how much information data carries about the parameter

## Jacobians and change of variables 

Show why Jacobians matter when transforming distributions.
	•	Simple change of variable in one dimension
	•	If $Y = g(X)$, then $f_Y(y) = f_X(x) \left| \frac{dx}{dy} \right|$
	•	Multivariate case: use Jacobian determinant
	•	Example: transforming from Cartesian to polar coordinates
	•	Example: bivariate normal to standard normal → Cholesky or whitening

Application: Why this matters in simulations and Bayesian inference (brief mention of Metropolis-Hastings, normalizing flows, etc.)

## Conclusions

- Calculus is not just background math—it’s the **engine** of statistical theory.
- Beyond undergraduate stats: everything from MLE, Bayesian posteriors, to machine learning involves gradients and Hessians.
 Suggest: for students who love math but unsure about statistics—you are the kind of person statistics needs.
